---
title: "Reproducible Research Project 1"
author: "Gloria Gabriel"
date: "1/8/2023"
output:
  md_document:
    variant: markdown_github
---

Reading the data set from the given url and unziping the dataset

```{r,echo=TRUE}
library("data.table")
library(ggplot2)

fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'), method = "curl")
unzip("repdata%2Fdata%2Factivity.zip",exdir = "data")
```
Converting the data into a csv format  that is suitable for my analysis
```{r,echo=TRUE}
activityDT <- data.table::fread(input = "data/activity.csv")
```

Obtaining the total number of steps taken per day.then displaying some observetions

```{r,echo=TRUE}
Total_Steps <- activityDT[, c(lapply(.SD, sum, na.rm = FALSE)), .SDcols = c("steps"), by = .(date)] 

head(Total_Steps, 10)

```

Creating a histogram on the total number of steps taken each day. The data created in the previous step will be used.

```{r,echo=TRUE}
ggplot(Total_Steps, aes(x = steps)) +
    geom_histogram(fill = "blue", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

Calculating the mean and median of the total number of steps taken in one day. The mean and median are displayed below.
```{r,echo=TRUE}
Total_Steps[, .(Mean_Steps = mean(steps, na.rm = TRUE), Median_Steps = median(steps, na.rm = TRUE))]
```
Creating a time series plot using ggplot of the 5-minutes interval and average number of steps taken.
```{r,echo=TRUE}
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval)]
ggplot(IntervalDT, aes(x = interval , y = steps)) + geom_line(color="blue", size=1) + labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")
```



the interval that contains the maximum number of steps can be obtained with the following codes
```{r,echo=TRUE}
IntervalDT[steps == max(steps), .(max_interval = interval)]
```
### missing values
to obtain the total number of missing values is done as follows, and obtain 2304 null entries
```{r,echo=TRUE}
activityDT[is.na(steps), .N ]
```
Using the median strategy to deal with the NA entries in the data set.

```{r,echo=TRUE}
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
```


creating a dataset called complete as a full dataset with no NA fillings.
```{r,echo=TRUE}
data.table::fwrite(x = activityDT, file = "data/completeData.csv", quote = FALSE)

###alternatively
complete<- activityDT
#checking if no empty entries
complete[is.na(steps), .N ]
```


```{r,echo=TRUE}
# total number of steps taken per day
Total_Steps <- complete[, c(lapply(.SD, sum)), .SDcols = c("steps"), by = .(date)] 

# mean and median total number of steps taken per day
Total_Steps[, .(Mean_Steps = mean(steps), Median_Steps = median(steps))]
```

```{r,echo=TRUE}
ggplot(Total_Steps, aes(x = steps)) + geom_histogram(fill = "blue", binwidth = 1000) + labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

```{r,echo=TRUE}
# Just recreating activityDT from scratch then making the new factor variable. (No need to, just want to be clear on what the entire process is.) 
activityDT <- data.table::fread(input = "data/activity.csv")
activityDT[, date := as.POSIXct(date, format = "%Y-%m-%d")]
activityDT[, `Day of Week`:= weekdays(x = date)]
activityDT[grepl(pattern = "Monday|Tuesday|Wednesday|Thursday|Friday", x = `Day of Week`), "weekday or weekend"] <- "weekday"
activityDT[grepl(pattern = "Saturday|Sunday", x = `Day of Week`), "weekday or weekend"] <- "weekend"
activityDT[, `weekday or weekend` := as.factor(`weekday or weekend`)]
head(activityDT, 10)
```

```{r,echo=TRUE}
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval, `weekday or weekend`)] 

ggplot(IntervalDT , aes(x = interval , y = steps, color=`weekday or weekend`)) + geom_line() + labs(title = "Avg. Daily Steps by Weektype", x = "Interval", y = "No. of Steps") + facet_wrap(~`weekday or weekend` , ncol = 1, nrow=2)
```

